Relational Parametricity for Algebraically Indexed Types

### 2012-05-17 paper/paper.tex
Move proofs of the lemmas to an appendix

\begin{proof}
  By induction on the derivation of $\Delta \vdash A \equiv B :
  \sortType$:
  \begin{description}
  \item[Case \TirName{TyEqUnit}] In this case, $A = B = \tyUnit$, so
    $\tySem{A} = \tySem{B}$.
  \item[Case \TirName{TyEqPrimIdx}] We have $A = \tyPrim{X}{e_1}$ and
    $B = \tyPrim{X}{e_2}$, for some $e_1$ and $e_2$ of the same
    sort. The erasure semantics ignores the index arguments, so
    $\tySem{A} = \tySem{\tyPrim{X}{e_1}} = \tyPrimSem{X} =
    \tySem{\tyPrim{X}{e_2}} = \tySem{B}$, as required.
  \item[Case \TirName{TyEqPrimNoIdx}] We have $A = \tyPrimNm{X}$ and
    $B = \tyPrimNm{X}$. So $\tySem{A} = \tyPrimSem{X} = \tySem{B}$, as
    required.
  \item[Case \TirName{TyEqArr}] Here, $A = A_1 \tyArr A_2$ and $B =
    B_1 \tyArr B_2$, and by the induction hypothesis we know that
    $\tySem{A_1} = \tySem{B_1}$ and $\tySem{A_2} = \tySem{B_2}$. Hence
    $\tySem{A} = \tySem{A_1 \tyArr A_2} = \tySem{A_1} \to \tySem{A_2}
    = \tySem{B_1} \to \tySem{B_2} = \tySem{B_1 \tyArr B_2} =
    \tySem{B}$.
  \item[Case \TirName{TyEqProd}] In this case, $A = A_1 \tyProduct
    A_2$ and $B = B_1 \tyProduct B_2$, and by the induction hypothesis
    we know that $\tySem{A_1} = \tySem{B_1}$ and $\tySem{A_2} =
    \tySem{B_2}$. Hence $\tySem{A} = \tySem{A_1 \tyProduct A_2} =
    \tySem{A_1} \times \tySem{A_2} = \tySem{B_1} \times \tySem{B_2} =
    \tySem{B_1 \tyProduct B_2} = \tySem{B}$.
  \item[Case \TirName{TyEqSum}] In this case, $A = A_1 + A_2$ and $B =
    B_1 + B_2$, and by the induction hypothesis we know that
    $\tySem{A_1} = \tySem{B_1}$ and $\tySem{A_2} = \tySem{B_2}$. Hence
    $\tySem{A} = \tySem{A_1 + A_2} = \tySem{A_1} + \tySem{A_2} =
    \tySem{B_1} + \tySem{B_2} = \tySem{B_1 + B_2} = \tySem{B}$.
  \item[Case \TirName{TyEqForall}] We have $A = \forall
    i\mathord:s.A'$ and $B = \forall i\mathord:s.B'$. By the induction
    hypothesis we know that $\tySem{A'} = \tySem{B'}$, and hence
    $\tySem{A} = \tySem{\forall i\mathord:s. A'} = \tySem{A'} =
    \tySem{B'} = \tySem{\forall i\mathord:s. B'} = \tySem{B}$.
  \end{description}
\end{proof}

@@

\begin{proof}
  By induction on the derivation of $\Delta \vdash A : \sortType$.
  \begin{description}
  \item[Case \TirName{TyUnit}] Directly from the definition of
    substitution on types of the form $\tyUnit$, we have
    $\tySem{\sigma^*\tyUnit} = \tySem{\tyUnit}$, as required.
  \item[Case \TirName{TyPrimIdx}] Since the index-erasure semantics
    ignores all index expessions, we have
    $\tySem{\sigma^*\tyPrim{X}{e}} = \tySem{\tyPrim{X}{\sigma^*e}} =
    \tyPrimSem{X} = \tySem{\tyPrim{X}{e}}$.
  \item[Case \TirName{TyPrimNoIdx}] Similar to the previous case.
  \item[Case \TirName{TyArr}] We have $A = A_1 \tyArr A_2$, and by the
    induction hypothesis, we know that $\tySem{\sigma^*A_1} =
    \tySem{A_1}$ and $\tySem{\sigma^*B_1} = \tySem{B_1}$. Thus, by the
    definition of substitution on arrow types, we have
    $\tySem{\sigma^*(A_1 \tyArr A_2)} = \tySem{\sigma^*A_1 \tyArr
      \sigma^*A_2} = \tySem{\sigma^*A_1} \to \tySem{\sigma^*A_2} =
    \tySem{A_1} \to \tySem{A_2} = \tySem{A_1 \tyArr A_2}$, as
    required.
  \item[Case \TirName{TyProd}] We have $A = A_1 \tyProduct A_2$, and
    by the induction hypothesis, we know that $\tySem{\sigma^*A_1} =
    \tySem{A_1}$ and $\tySem{\sigma^*A_2} = \tySem{A_2}$. Thus, by the
    definition of substitution on product types, we have
    $\tySem{\sigma^*(A_1 \tyProduct A_2)} = \tySem{\sigma^*A_1
      \tyProduct \sigma^*A_2} = \tySem{\sigma^*A_1} \times
    \tySem{\sigma^*A_2} = \tySem{A_1} \times \tySem{A_2} = \tySem{A_1
      \tyProduct A_2}$, as required.
  \item[Case \TirName{TySum}] We have $A = A_1 + A_2$, and by the
    induction hypothesis, we know that $\tySem{\sigma^*A_1} =
    \tySem{A_1}$ and $\tySem{\sigma^*A_2} = \tySem{A_2}$. Thus, by the
    definition of substitution on sum types, we have
    $\tySem{\sigma^*(A_1 + A_2)} = \tySem{\sigma^*A_1 + \sigma^*A_2} =
    \tySem{\sigma^*A_1} + \tySem{\sigma^*A_2} = \tySem{A_1} +
    \tySem{A_2} = \tySem{A_1 + A_2}$, as required.
  \item[Case \TirName{TyForall}] We have $A = \forall i\mathord:s.A'$,
    and by the induction hypothesis we know that, for all $\sigma'$,
    $\tySem{\sigma'^*A'} = \tySem{A'}$. Therefore, from the definition
    of substitution for $\forall$-types, we have
    $\tySem{\sigma^*\forall i\mathord:s.A'} = \tySem{\forall
      i\mathord:s.\sigma_s^{*}A'} = \tySem{\sigma_s^{*}A'} = \tySem{A'}
    = \tySem{\forall i\mathord:s.A'}$, where we have relied on the
    fact that the index-erasure semantics of types ignores the index
    quantification.
  \end{description}
\end{proof}

@@

\begin{proof}
  By induction on the derivation of $\Delta \vdash A \equiv B :
  \sortType$.
  \begin{description}
  \item[Case \TirName{TyEqUnit}] In this case $A = B = \tyUnit$, so
    the statement is trivially statisfied.
  \item[Case \TirName{TyEqPrimIdx}] In this case we have $A =
    \tyPrim{X}{e_1}$ and $B = \tyPrim{X}{e_2}$, with $\Delta \vdash
    e_1 \equiv e_2 : s$. The index expressions $e_1$ and $e_2$ are in
    the same equivalence class, so by the definition of relation
    environments we have that $\rho_{\tyPrimNm{X}}(e_1) =
    \rho_{\tyPrimNm{X}}(e_2)$ and hence
    $\rsem{\tyPrim{X}{e_1}}{\relEnv{E}}\rho =
    \rsem{\tyPrim{X}{e_2}}{\relEnv{E}}\rho$.
  \item[Case \TirName{TyEqPrimNoIdx}] This case is similar to the case
    for \TirName{TyEqUnit}.
  \item[Case \TirName{TyEqArr}] We have $A = A_1 \tyArr A_2$ and $B =
    B_1 \tyArr B_2$, with $\Delta \vdash A_1 \equiv B_1 : \sortType$
    and $\Delta \vdash A_2 \equiv B_2 : \sortType$. By the induction
    hypothesis, we know that $\rsem{A_1}{\relEnv{E}}\rho =
    \rsem{B_1}{\relEnv{E}}\rho$ and $\rsem{A_2}{\relEnv{E}}\rho =
    \rsem{B_2}{\relEnv{E}}\rho$. Thus we may reason as follows:
    $\rsem{A_1 \tyArr A_2}{\relEnv{E}}\rho =
    \rsem{A_1}{\relEnv{E}}\rho \relArrow \rsem{A_2}{\relEnv{E}}\rho =
    \rsem{B_1}{\relEnv{E}}\rho \relArrow \rsem{B_2}{\relEnv{E}}\rho =
    \rsem{B_1 \tyArr B_2}{\relEnv{E}}\rho$, as required.
  \item[Case \TirName{TyEqProd}] We have $A = A_1 \tyProduct A_2$ and
    $B = B_1 \tyProduct B_2$, with $\Delta \vdash A_1 \equiv B_1 :
    \sortType$ and $\Delta \vdash A_2 \equiv B_2 : \sortType$. By the
    induction hypothesis, we know that $\rsem{A_1}{\relEnv{E}}\rho =
    \rsem{B_1}{\relEnv{E}}\rho$ and $\rsem{A_2}{\relEnv{E}}\rho =
    \rsem{B_2}{\relEnv{E}}\rho$. Therefore, we can reason as follows:
    $\rsem{A_1 \tyProduct A_2}{\relEnv{E}}\rho =
    \rsem{A_1}{\relEnv{E}}\rho \relTimes \rsem{A_2}{\relEnv{E}}\rho =
    \rsem{B_1}{\relEnv{E}}\rho \relTimes \rsem{B_2}{\relEnv{E}}\rho =
    \rsem{B_1 \tyProduct B_2}{\relEnv{E}}\rho$, as required.
  \item[Case \TirName{TyEqSum}] We have $A = A_1 + A_2$ and $B = B_1 +
    B_2$, with $\Delta \vdash A_1 \equiv B_1 : \sortType$ and $\Delta
    \vdash A_2 \equiv B_2 : \sortType$. By the induction hypothesis,
    we know that $\rsem{A_1}{\relEnv{E}}\rho =
    \rsem{B_1}{\relEnv{E}}\rho$ and $\rsem{A_2}{\relEnv{E}}\rho =
    \rsem{B_2}{\relEnv{E}}\rho$. Therefore, we can reason as follows:
    $\rsem{A_1 + A_2}{\relEnv{E}}\rho = \rsem{A_1}{\relEnv{E}}\rho
    \relSum \rsem{A_2}{\relEnv{E}}\rho = \rsem{B_1}{\relEnv{E}}\rho
    \relSum \rsem{B_2}{\relEnv{E}}\rho = \rsem{B_1 +
      B_2}{\relEnv{E}}\rho$, as required.
  \item[Case \TirName{TyEqForall}] We have $A = \forall
    i\mathord:s. A'$ and $B = \forall i\mathord:s.B'$, with $\Delta, i
    : s \vdash A' \equiv B' : \sortType$. By the induction hypothesis,
    we know that for all $\rho'$, $\rsem{A'}{\relEnv{E}}\rho' =
    \rsem{B'}\rho'$. We reason as follows:
    \begin{eqnarray*}
      &    & (x_1,x_2) \in \rsem{\forall i\mathord:s. A'}{\relEnv{E}}\rho \\
      &\iff& \forall \rho' \in \extends{\rho}{i:s}.\ (x_1,x_2) \in \rsem{A'}{\relEnv{E}}\rho' \\
      &\iff& \forall \rho' \in \extends{\rho}{i:s}.\ (x_1,x_2) \in \rsem{B'}{\relEnv{E}}\rho' \\
      &\iff& (x_1,x_2) \in \rsem{\forall i\mathord:s. B'}{\relEnv{E}}\rho
    \end{eqnarray*}
    Thus, by extensionality, we have $\rsem{\forall
      i\mathord:s.A'}\rho \equiv \rsem{\forall
      i\mathord:s.B'}{\relEnv{E}}\rho$, as required.
  \end{description}
\end{proof}

@@

\begin{proof}

>>>>

@@

@@

@@

\appendix

\section{Details of proofs}

\begin{proof}\textbf{[Of \lemref{lem:tyeq-erasure}]}
  By induction on the derivation of $\Delta \vdash A \equiv B :
  \sortType$:
  \begin{description}
  \item[Case \TirName{TyEqUnit}] In this case, $A = B = \tyUnit$, so
    $\tySem{A} = \tySem{B}$.
  \item[Case \TirName{TyEqPrimIdx}] We have $A = \tyPrim{X}{e_1}$ and
    $B = \tyPrim{X}{e_2}$, for some $e_1$ and $e_2$ of the same
    sort. The erasure semantics ignores the index arguments, so
    $\tySem{A} = \tySem{\tyPrim{X}{e_1}} = \tyPrimSem{X} =
    \tySem{\tyPrim{X}{e_2}} = \tySem{B}$, as required.
  \item[Case \TirName{TyEqPrimNoIdx}] We have $A = \tyPrimNm{X}$ and
    $B = \tyPrimNm{X}$. So $\tySem{A} = \tyPrimSem{X} = \tySem{B}$, as
    required.
  \item[Case \TirName{TyEqArr}] Here, $A = A_1 \tyArr A_2$ and $B =
    B_1 \tyArr B_2$, and by the induction hypothesis we know that
    $\tySem{A_1} = \tySem{B_1}$ and $\tySem{A_2} = \tySem{B_2}$. Hence
    $\tySem{A} = \tySem{A_1 \tyArr A_2} = \tySem{A_1} \to \tySem{A_2}
    = \tySem{B_1} \to \tySem{B_2} = \tySem{B_1 \tyArr B_2} =
    \tySem{B}$.
  \item[Case \TirName{TyEqProd}] In this case, $A = A_1 \tyProduct
    A_2$ and $B = B_1 \tyProduct B_2$, and by the induction hypothesis
    we know that $\tySem{A_1} = \tySem{B_1}$ and $\tySem{A_2} =
    \tySem{B_2}$. Hence $\tySem{A} = \tySem{A_1 \tyProduct A_2} =
    \tySem{A_1} \times \tySem{A_2} = \tySem{B_1} \times \tySem{B_2} =
    \tySem{B_1 \tyProduct B_2} = \tySem{B}$.
  \item[Case \TirName{TyEqSum}] In this case, $A = A_1 + A_2$ and $B =
    B_1 + B_2$, and by the induction hypothesis we know that
    $\tySem{A_1} = \tySem{B_1}$ and $\tySem{A_2} = \tySem{B_2}$. Hence
    $\tySem{A} = \tySem{A_1 + A_2} = \tySem{A_1} + \tySem{A_2} =
    \tySem{B_1} + \tySem{B_2} = \tySem{B_1 + B_2} = \tySem{B}$.
  \item[Case \TirName{TyEqForall}] We have $A = \forall
    i\mathord:s.A'$ and $B = \forall i\mathord:s.B'$. By the induction
    hypothesis we know that $\tySem{A'} = \tySem{B'}$, and hence
    $\tySem{A} = \tySem{\forall i\mathord:s. A'} = \tySem{A'} =
    \tySem{B'} = \tySem{\forall i\mathord:s. B'} = \tySem{B}$.
  \end{description}
\end{proof}

\begin{proof}\textbf{[Of \lemref{lem:tysubst-erasure}]}
  By induction on the derivation of $\Delta \vdash A : \sortType$.
  \begin{description}
  \item[Case \TirName{TyUnit}] Directly from the definition of
    substitution on types of the form $\tyUnit$, we have
    $\tySem{\sigma^*\tyUnit} = \tySem{\tyUnit}$, as required.
  \item[Case \TirName{TyPrimIdx}] Since the index-erasure semantics
    ignores all index expessions, we have
    $\tySem{\sigma^*\tyPrim{X}{e}} = \tySem{\tyPrim{X}{\sigma^*e}} =
    \tyPrimSem{X} = \tySem{\tyPrim{X}{e}}$.
  \item[Case \TirName{TyPrimNoIdx}] Similar to the previous case.
  \item[Case \TirName{TyArr}] We have $A = A_1 \tyArr A_2$, and by the
    induction hypothesis, we know that $\tySem{\sigma^*A_1} =
    \tySem{A_1}$ and $\tySem{\sigma^*B_1} = \tySem{B_1}$. Thus, by the
    definition of substitution on arrow types, we have
    $\tySem{\sigma^*(A_1 \tyArr A_2)} = \tySem{\sigma^*A_1 \tyArr
      \sigma^*A_2} = \tySem{\sigma^*A_1} \to \tySem{\sigma^*A_2} =
    \tySem{A_1} \to \tySem{A_2} = \tySem{A_1 \tyArr A_2}$, as
    required.
  \item[Case \TirName{TyProd}] We have $A = A_1 \tyProduct A_2$, and
    by the induction hypothesis, we know that $\tySem{\sigma^*A_1} =
    \tySem{A_1}$ and $\tySem{\sigma^*A_2} = \tySem{A_2}$. Thus, by the
    definition of substitution on product types, we have
    $\tySem{\sigma^*(A_1 \tyProduct A_2)} = \tySem{\sigma^*A_1
      \tyProduct \sigma^*A_2} = \tySem{\sigma^*A_1} \times
    \tySem{\sigma^*A_2} = \tySem{A_1} \times \tySem{A_2} = \tySem{A_1
      \tyProduct A_2}$, as required.
  \item[Case \TirName{TySum}] We have $A = A_1 + A_2$, and by the
    induction hypothesis, we know that $\tySem{\sigma^*A_1} =
    \tySem{A_1}$ and $\tySem{\sigma^*A_2} = \tySem{A_2}$. Thus, by the
    definition of substitution on sum types, we have
    $\tySem{\sigma^*(A_1 + A_2)} = \tySem{\sigma^*A_1 + \sigma^*A_2} =
    \tySem{\sigma^*A_1} + \tySem{\sigma^*A_2} = \tySem{A_1} +
    \tySem{A_2} = \tySem{A_1 + A_2}$, as required.
  \item[Case \TirName{TyForall}] We have $A = \forall i\mathord:s.A'$,
    and by the induction hypothesis we know that, for all $\sigma'$,
    $\tySem{\sigma'^*A'} = \tySem{A'}$. Therefore, from the definition
    of substitution for $\forall$-types, we have
    $\tySem{\sigma^*\forall i\mathord:s.A'} = \tySem{\forall
      i\mathord:s.\sigma_s^{*}A'} = \tySem{\sigma_s^{*}A'} = \tySem{A'}
    = \tySem{\forall i\mathord:s.A'}$, where we have relied on the
    fact that the index-erasure semantics of types ignores the index
    quantification.
  \end{description}
\end{proof}

\begin{proof}\textbf{[Of \lemref{lem:tyeq-rel}]}
  By induction on the derivation of $\Delta \vdash A \equiv B :
  \sortType$.
  \begin{description}
  \item[Case \TirName{TyEqUnit}] In this case $A = B = \tyUnit$, so
    the statement is trivially statisfied.
  \item[Case \TirName{TyEqPrimIdx}] In this case we have $A =
    \tyPrim{X}{e_1}$ and $B = \tyPrim{X}{e_2}$, with $\Delta \vdash
    e_1 \equiv e_2 : s$. The index expressions $e_1$ and $e_2$ are in
    the same equivalence class, so by the definition of relation
    environments we have that $\rho_{\tyPrimNm{X}}(e_1) =
    \rho_{\tyPrimNm{X}}(e_2)$ and hence
    $\rsem{\tyPrim{X}{e_1}}{\relEnv{E}}\rho =
    \rsem{\tyPrim{X}{e_2}}{\relEnv{E}}\rho$.
  \item[Case \TirName{TyEqPrimNoIdx}] This case is similar to the case
    for \TirName{TyEqUnit}.
  \item[Case \TirName{TyEqArr}] We have $A = A_1 \tyArr A_2$ and $B =
    B_1 \tyArr B_2$, with $\Delta \vdash A_1 \equiv B_1 : \sortType$
    and $\Delta \vdash A_2 \equiv B_2 : \sortType$. By the induction
    hypothesis, we know that $\rsem{A_1}{\relEnv{E}}\rho =
    \rsem{B_1}{\relEnv{E}}\rho$ and $\rsem{A_2}{\relEnv{E}}\rho =
    \rsem{B_2}{\relEnv{E}}\rho$. Thus we may reason as follows:
    $\rsem{A_1 \tyArr A_2}{\relEnv{E}}\rho =
    \rsem{A_1}{\relEnv{E}}\rho \relArrow \rsem{A_2}{\relEnv{E}}\rho =
    \rsem{B_1}{\relEnv{E}}\rho \relArrow \rsem{B_2}{\relEnv{E}}\rho =
    \rsem{B_1 \tyArr B_2}{\relEnv{E}}\rho$, as required.
  \item[Case \TirName{TyEqProd}] We have $A = A_1 \tyProduct A_2$ and
    $B = B_1 \tyProduct B_2$, with $\Delta \vdash A_1 \equiv B_1 :
    \sortType$ and $\Delta \vdash A_2 \equiv B_2 : \sortType$. By the
    induction hypothesis, we know that $\rsem{A_1}{\relEnv{E}}\rho =
    \rsem{B_1}{\relEnv{E}}\rho$ and $\rsem{A_2}{\relEnv{E}}\rho =
    \rsem{B_2}{\relEnv{E}}\rho$. Therefore, we can reason as follows:
    $\rsem{A_1 \tyProduct A_2}{\relEnv{E}}\rho =
    \rsem{A_1}{\relEnv{E}}\rho \relTimes \rsem{A_2}{\relEnv{E}}\rho =
    \rsem{B_1}{\relEnv{E}}\rho \relTimes \rsem{B_2}{\relEnv{E}}\rho =
    \rsem{B_1 \tyProduct B_2}{\relEnv{E}}\rho$, as required.
  \item[Case \TirName{TyEqSum}] We have $A = A_1 + A_2$ and $B = B_1 +
    B_2$, with $\Delta \vdash A_1 \equiv B_1 : \sortType$ and $\Delta
    \vdash A_2 \equiv B_2 : \sortType$. By the induction hypothesis,
    we know that $\rsem{A_1}{\relEnv{E}}\rho =
    \rsem{B_1}{\relEnv{E}}\rho$ and $\rsem{A_2}{\relEnv{E}}\rho =
    \rsem{B_2}{\relEnv{E}}\rho$. Therefore, we can reason as follows:
    $\rsem{A_1 + A_2}{\relEnv{E}}\rho = \rsem{A_1}{\relEnv{E}}\rho
    \relSum \rsem{A_2}{\relEnv{E}}\rho = \rsem{B_1}{\relEnv{E}}\rho
    \relSum \rsem{B_2}{\relEnv{E}}\rho = \rsem{B_1 +
      B_2}{\relEnv{E}}\rho$, as required.
  \item[Case \TirName{TyEqForall}] We have $A = \forall
    i\mathord:s. A'$ and $B = \forall i\mathord:s.B'$, with $\Delta, i
    : s \vdash A' \equiv B' : \sortType$. By the induction hypothesis,
    we know that for all $\rho'$, $\rsem{A'}{\relEnv{E}}\rho' =
    \rsem{B'}\rho'$. We reason as follows:
    \begin{eqnarray*}
      &    & (x_1,x_2) \in \rsem{\forall i\mathord:s. A'}{\relEnv{E}}\rho \\
      &\iff& \forall \rho' \in \extends{\rho}{i:s}.\ (x_1,x_2) \in \rsem{A'}{\relEnv{E}}\rho' \\
      &\iff& \forall \rho' \in \extends{\rho}{i:s}.\ (x_1,x_2) \in \rsem{B'}{\relEnv{E}}\rho' \\
      &\iff& (x_1,x_2) \in \rsem{\forall i\mathord:s. B'}{\relEnv{E}}\rho
    \end{eqnarray*}
    Thus, by extensionality, we have $\rsem{\forall
      i\mathord:s.A'}\rho \equiv \rsem{\forall
      i\mathord:s.B'}{\relEnv{E}}\rho$, as required.
  \end{description}
\end{proof}

\begin{proof}\textbf{[Of \lemref{lem:tysubst-rel}]}
